An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-15 thesis/introduction-and-background.tex
update sections and add content to thesis

\chapter{Preliminaries}

\section{Graphics Processing Units (GPUs)}

\subsection{Motivation}

Most numerical approaches for modeling systems and phenomena
approximate continuous quantities as discrete quantities.
The larger the number of discrete entities used,
the better the approximation.
In CFD, the physical system (fluid)
is represented as a set of discrete points (finite difference methods),
or discrete cells (finite volume method),
and this discrete representation is termed as a \emph{mesh}.
In CFD applications
a mesh containing several points is used
to represent the underlying system,
and numerical simulations are performed
to study how physical quantities
change with time at these mesh points (or cells).
Typically, a \emph{time stepping} scheme is used to study this evolution,
wherein the simulation proceeds as a series of \emph{time steps}.
For each time step, the quantities at each mesh point at
\emph{updated} from quantities computed at the previous time step.

In the direct numerical simulation (DNS) of fluid systems,
the aim is to resolve the smallest features of the flow.
This requires the use of extremely fine meshes.
At each of the mesh points,
data about the flow must be stored
(for instance, the pressure, temperature and velocities at that point).
The large amount of data storage required places
restrictions on the problem sizes that can be accommodated
on a single workstation/PC.
Further,
the numerical simulation requires a large amount of
computations with this data.
In general, the computations are of a repetitive nature:
the same computations are performed repeatedly
for the different mesh points,
and for successive time steps.

Thus, CFD is typically a computationally expensive endeavour,
and it has long been of interest to use
\emph{parallel processing} to
perform numerical simulations of fluid flows.
This involves distributing computing tasks
among multiple independent processors,
which can run concurrently.
Of course, this requires that the computations be \emph{independent}
of each other.
Fortunately, the computational patterns that arise in CFD
typically satisfy this requirement:
the computations at different mesh points are largely independent
of each other and can be done in parallel.

The traditional approach to parallel computing has been
to distribute parallel tasks among CPU cores.
The cores may all have access to a common memory space
(shared memory systems),
or may have distinct memory spaces
(distributed memory systems).
Much more common than purely distributed systems are
\emph{hybrid} systems wherein the tasks may be distributed
among so-called \emph{nodes},
which are themselves shared-memory systems.
Again, the computational workhorse in all these systems has traditionally
been the CPU core.

A more recent parallel system that has seen
wide adoption
is the Graphics Processing Unit (GPU).
GPUs were developed primarily for performing
graphics computations, i.e.,
calculations with pixel data.
This is a \emph{highly} parallelizable task;
pixels can be processed independently and in parallel.
Thus, the GPU evolved as a system
to perform hundreds of computations concurrently,
and it should be no surprise that it soon
gained attention from the scientific computation community.
Now, several scientific software packages have been
written for GPUs,
and the worlds largest supercomputers employ GPUs
as accelerators.

@@

The approximation of higher order derivatives generally requires
inclusion of a larger number of points in the
finite difference approximation
(referred to as the finite difference \emph{stencil}).
For a given order of derivative,
finite difference approximation of arbitrary stencil widths
may be derived,
with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.
In most applications,
large stencil widths are undesirable.
This is because:

\begin{enumerate}
    \item The arithmetic intensity increases, i.e.,
        a larger number of computations must be performed per grid point.
    \item When the solution is done in parallel,
        the amount of boundary information that must be exchanged
        between processors increases.
\end{enumerate}

To overcome the problems associated with larger stencil widths,
another class of finite difference schemes,
known as \emph{compact} schemes may be used.
First proposed by Sanjeeva Lele ~\cite{lele1992compact},
compact schemes express the derivative at a point $i$
in terms of the derivatives at neighbouring points,
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:

\begin{equation}
\begin{split}
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints.
We note that for equations of the form in Eq. \ref{eqn:general-compact},
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
and $A$ is, in general, a \emph{banded} matrix.

The simplest classes of banded matrix include
diagonal, tridiagonal and pentadiagonal matrices.
When $A$ is diagonal ($\alpha$ = $\beta$ = $\hdots$ = 0),
the scheme is explicit and straightforward to evaluate:
the derivatives are simply computed pointwise
(in series or in parallel),
by application of the right-hand side stencil at each point.
When $A$ is tridiagonal ($\alpha \neq 0$, $\beta$ = $\hdots$ = 0),
the evaluation of the derivatives requires the solution
of the tridiagonal system, which yields the derivatives
at all points simultaneously.
The classical Thomas algorithm
also referred to as the Tridiagonal Matrix Algorithm,
is widely employed for this purpose.
The algorithm is straightforward to implement,
and stable for diagonally dominant matrices ~\cite{numericalrecipes}.

>>>>

\chapter{Introduction and Background}

\section{Motivation}

In the direct numerical simulation (DNS) of fluid systems,
the aim is to resolve the smallest features of the flow
without resorting to any modeling of the turbulence.
This involves the use of
extremely fine computational grids to
discretize the flow domain,
and some numerical method for solving the
flow equations on this grid.
By far the most popular numerical approach for
solving the flow equations
is the \emph{finite difference method}.
The finite difference method approximates the
spatial and temporal derivatives appearing in the
partial differential equations that describe the flow
using finite difference schemes.
\emph{Compact finite difference schemes} are a class
of finite difference schemes that have found widespread
adoption in DNS codes
for their high order of accuracy.
The evaluation of compact finite differences
requires the repeated solution of \emph{banded} linear systems,
making them fairly complex and expensive computationally.

As mentioned, DNS involves the use of extremely fine grids.
At each of the grid points,
various data about the flow must be stored---for
example, the geometric coordinates (x, y and z),
pressure, temperature and velocities at that point.
The amount of memory required to store this data
for even small problem sizes
can quickly exceed the capacity of modern workstations/PCs.
Thus, a \emph{distributed memory} parallel system
is generally required for performing DNS.
Here, the traditional approach has been
to distribute parallel tasks among individual CPU cores,
or groups of CPU cores
that share common memory spaces.
In the latter case,
each group of cores constitutesa \emph{shared memory} system,
and the overall system is referred to as a \emph{hybrid system}.

The workhorse for computation in the above described
parallel systems is the CPU core,
however, more recently,
Graphics Processing Units (GPUs)
are being used for performing intensive calculations.
GPUs, while themselves being highly parallel processors,
can also function as accelerators in distributed memory systems
(GPU clusters).
However,
applications that exploit such systems
need careful redesign of algorithms---and
sometimes, substantial changes to code---to
see significant performance improvements.
This is because the CPU and GPU have very different architectures,
and any na\"{\i}ve ``parallelization'' of algorithms
designed for the CPU
is likely not take full advantage of the GPU's
memory heirarchy and compute ability.

The objective of this work is to develop an
approach for evaluating
compact finite differences
on multiple GPUs in a distributed system.
This is of interest because
the evaluation of spatial derivatives using
compact finite difference schemes is one of the most
expensive tasks in DNS.
But perhaps more significantly,
it encompasses several computational patterns
such as pointwise updates, stencil evaluations,
and solutions of distributed tridiagonal systems.
Efficient implementation of these computational patterns
on GPUs is of interest in other areas of CFD,
and scientific computation in general.

@@

The approximation of higher order derivatives generally requires
inclusion of a larger number of points in the
finite difference approximation
(referred to as the finite difference \emph{stencil}).
For a given order of derivative,
finite difference approximation of arbitrary stencil widths
may be derived,
with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.

\subsection{Wavenumber analysis of finite difference methods}

For DNS applications, the relevant measure of accuracy of a
finite difference scheme is obtained from the so-called
\emph{modified wavenumber} approach.
Here, we test the scheme's ability to
accurately estimate the derivative of sinusoidal functions
with increasing wavenumbers (frequencies),
given a specified grid size. 
We expect that the approximation of the derivative
becomes more difficult with increasing wavenumbers,
as the function value varies more rapidly.
The wavenumber of the approximate derivative
as given by the finite difference scheme is compared with
the wavenumber of the exact derivative.
The result for different schemes is shown
in Fig. \ref{fig:modified-wavenumbers}.

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{fig/modified-wavenumbers.eps}
\end{center}
\caption{Modified wavenumbers for three different schemes
    with the same stencil width.
    The compact schemes better estimate the derivative
    for higher wavenumbers.}
\label{fig:modified-wavenumbers}
\end{figure}

The ability of a finite difference scheme
to accomodate large wavenumbers
is extremely relevant in
computational fluid dynamics applications
~\cite{KravchenkoEffNumErr}.
In DNS, the scheme
\emph{must} capture the
rapidly varying characteristics of the flow
associated with the turbulence.
For this purpose,
higher order explicit schemes
may be considered.
While they are traightforward to compute,
they are associated with large stencil widths.
In most applications,
large stencil widths are undesirable.
This is because
the arithmetic intensity increases with stencil size, i.e.,
a larger number of computations must be performed per grid point.
Further,
the amount of boundary information that must be exchanged
between parallel processes increases,
which can seriously affect overall performance.

From \ref{fig:modified-wavenumbers},
it is clear that the compact schemes are better
able to compute derivatives for larger wavenumbers
for a given stencil width.
Thus, they are widely used in
CFD for evaluation of spatial derivatives.

\subsection{General form of compact finite difference schemes}
\label{subsec:gen-form-compact}

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{lele1992compact}.
We note that for equations of the form in Eq. \ref{eqn:general-compact},
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq.\ref{eqn:general-compact}],
and $A$ is, in general, a \emph{banded} matrix.

\subsection{Boundary conditions}

One of the advantages of the compact finite difference approach
is that it accommodates \emph{non-periodic} boundary conditions.
This is in contrast to other methods used in DNS,
such as the \emph{spectral} methods.
We note that the Eq. (\ref{eqn:general-compact})
cannot be applied near the boundary points.
At the boundaries,
\emph{non-centered} or \emph{one-sided}
finite difference approximations are required.
Some considerations are made in choosing
these approximations:
firstly, the bandwidth of the resulting banded matrix
must be preserved.
Secondly,
the width of the boundary stencils
must be lower than the interior stencils,
as higher order boundary stencils are unstable
~\cite{kennedy1994several}.
In general, boundary equations for the first derivative
are of the following form ~\cite{lele1992compact}:

\begin{equation}
    f_1^{\prime} + \alpha f_2^{\prime} = \
        \frac{1}{h}(af_1 + bf_2 + cf_3 + df_4)
\label{eqn:boundary-compact}
\end{equation}

\subsection{Tridiagonal compact schemes}

As mentioned in Sec. \ref{subsec:gen-form-compact},
compact finite difference schemes lead to
\emph{banded} linear systems.
The simplest classes of banded matrix include
diagonal, tridiagonal and pentadiagonal matrices.
When $\alpha$ = $\beta$ = $\hdots$ = 0,
$A$ is a $\emph{diagonal}$ matrix.
In this case, the
scheme is \emph{explicit}
and the derivatives are straightforward to evaluate:
this simply involves the
application of the right-hand side stencil at each point.
When $\alpha \neq 0, \beta = \hdots$ = 0,
$A$ is \emph{tridiagonal}.
The evaluation of the derivatives requires
the solution of the resulting tridiagonal system.
The classical Thomas algorithm---also
referred to as the Tridiagonal Matrix Algorithm---is
widely employed for this purpose.
The algorithm is straightforward to implement,
and stable for diagonally dominant matrices ~\cite{numericalrecipes}.
When $\alpha \neq 0$ and $\beta \neq  0$,
pentadiagonal and higher-bandwidth systems arise.
These can be more expensive and difficult to evaluate
than tridiagonal systems
using a direct method.

This work will focus on
compact finite difference schemes that lead
to tridiagonal systems.
An example of such a scheme is obtained
by substituting 
$\beta = 0$, $\alpha = \frac{1}{4}$ and $a = \frac{3}{4}$
in Eq. \ref{eqn:general-compact}
(all other coefficients are set to zero).
This leads to a fourth-order accurate
tridiagonal compact finite difference scheme,
known also as the Pad\'{e} scheme.
At the boundaries,
the following \emph{third-order} accurate
approximations are used:

\begin{align}
    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx} \\
    f^{\prime}_{n} + 2f^{\prime}_{n-1}
    =
    \frac{5f_{n} - 4f_{n-2} -  f_{n-1}}{dx}
\end{align}
%
The resulting tridiagonal system is then:

\begin{equation}
 \label{eqn:compact-tridiagonal-system}
 \begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
  \begin{bmatrix}
      f^{\prime}_1 \\

      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     \frac{-5f_1 + 4f_2 + f_3}{2}\\
     \frac{3(f_{3} - f_{1})}{4}\\
     \frac{3(f_{4} - f_{2})}{4}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     \frac{3(f_{n} - f_{n-2})}{4}\\
     \frac{5f_{n} - 4f_{n-1} - f_{n-2}}{2}
  \end{bmatrix}
\end{equation}

Similar tridiagonal schemes are available
to evaluate higher order derivatives.
Compact schemes for
second, third and fourth derivatives
are derived in ~\cite{lele1992compact}.

\section{Graphics processing units}

\subsection{Introduction}
\subsection{Memory model}
\subsubsection{Global memory}
\subsubsection{Shared memory}
\subsubsection{Registers and local memory}
\subsection{CUDA Programming model}
\subsubsection{Kernels}
\subsubsection{Thread organization}

\section{Tridiagonal solvers}
\subsection{Thomas algorithm}
\subsection{Cyclic reduction}
\subsection{Other algorithms}
